Many claims of group polarization are plausibly false due to inappropriate statistics

### 2022-11-20 GroupPolarizationStatmod/paper/ms.tex
some progress writing up model and outlining results/discussion

\documentclass[11pt, letterpaper]{article}
\usepackage{marvosym}
\usepackage{setspace}
\usepackage{longtable}
% DOCUMENT LAYOUT
% \usepackage{geometry} 
\usepackage[margin=1.0in]{geometry}
\usepackage{lineno}
\usepackage[usenames,dvipsnames]{xcolor}
% \geometry{letterpaper, textwidth=5.5in, textheight=8.5in, marginparsep=7pt, marginparwidth=.6in}
% \setlength\parindent{0in}

@@

@@

Our model represnts the generative process of finding different forms of group consensus
in group polarization experiments, and the process of measuring that consensus using
an ordinal measurement scale, e.g.\ a Likert-type scale. This model considers two
relevant forms of consensus: \emph{simple consensus} and \emph{group polarization}.
Either process is \emph{consensus} since opinion variance among the group 
decreases as individuals come to agree more than they did previously. 
Group polarization experiments are designed to foster group consensus
it is not necessary to consider clustering, negative influence, or other
processes known to generally occur when opinions within a group significantly
diverge~\cite{Turner2018}.

We collected group polarization data from ten empirical studies with sixty constituent
experimental treatments across all studies, specifically the opinion shift and pre- and
post-deliberation mean opinions, and the number of participants
in each treatment (\citeA{Burnstein1975} did not report pre- and post-deliberation
means, only the shift in means). We did not include experimental treatments that did not 
yield a positive detection of group polarization since we are only estimating
the rate of plausible false positive detections of group polarization. 
We then created new data that indicated whether a reported.

In two cases, reported opinion and shift values were converted from an ordinal
scale to a percentage scale, specifically \citeA{Friedkin1999a} and \citeA{Krizan2007}.
The authors in these papers did report the original ordinal measurement scale,
so we converted the means back to the orignial measurement scale.
In a few cases from \citeA{Friedkin1999a} we did not need to apply our 
model to the reported data because we could determine a reported opinion shift was 
\emph{prima facie} false since reported opinion shift magnitudes
were smaller than the standard deviation in shifts (see Table 1, p. 868, \emph{ibid}),
despite Friedkin's claim that regressions (not clearly defined) on all shifts 
yielded $p < 0.05$~\cite{Kruschke2018c}.

\subsection{Latent opinions, ordinal measurements}

Our analysis relies on two important formal concepts: latent opinions and their 
measurement on an ordinal scale. \emph{Latent opinions} are unobservable psychological,
continuous-valued variables that are somehow represented and stored in the
brain-body. 
When researchers use continuous statistical models to infer 
that group polarization occurred (which all ten considered here do), 
they are implicitly assuming that latent opinions are themselves continuous.
We can only measure behaviors in group polarization experiments,
specifically the behavior of participants marking one ordinal value or another
to indicate their opinion---a process that is not well understood in itself,
but which is unimportant for our analyses. \emph{Ordinal measurements} (e.g.\
a Likert or other discrete-valued opinion scale) of latent opinions force
participants to appropriately bin their opinion in one of several categories
representing the valence and extremity of their opinion, e.g., on a
7-point Likert scale -3 could indcate
``strongly disagree'', +3 could be ``strongly agree'', 0 would be neutral, and 
intermediate values are less extreme opinions of either valence.

@@

\subsection{Simple consensus and group polarization are statistical patterns}

\subsection{Analytical approach}

The goal of this study is to understand if published detections of group polarization are
plausibly false.

Therefore we designed a series of tests to understand if, and to what degree,
published results are false positive detections of group polarization that
are at least as well described by a simple-consensus model.

\begin{equation}
  \int_{-\infty}^\infty x^2 dx \neq \pi
\end{equation}
\noindent

\subsubsection{Identifying parameters that plausibly generate false detections of group polarization}

@@

\subsubsection{Bayesian ordered-probit models of generated simple-consensus data}

@@

We found that in X\% of generated simple consensus data led to false positive
detections

When simulated data reliably result in false positive $t$-test results (where
the $t$-test $p \le 0.1$) this strengthens the case that reported detections
of a false group polarization detection.

\spacing{1.0}
% \input{t-test-table.tex}
\input{t-test-table-custom.tex}
\spacing{1.5}

@@

This problem is made all the worse by the lack of a consistent experimental design for group
polarization experiments. This makes it impossible to systematically account for
additional sources of vairance known to lead to (orthogonal) over-estimates
of effect sizes~\cite{Yarkoni2021}. This is due to the rush to explain why 
group polarization occurs using weak theory~\cite{Cartwright1971,Cartwright1973} and makes
group polarization meta-analyses (e.g. Brown, 1986; and Isenberg, 1986)\nocite{Brown1986,Isenberg1986}
unreliable~\cite{Meehl1990,Meehl1997}.

>>>>

\documentclass[11pt, letterpaper]{article}
\usepackage{marvosym}
\usepackage{setspace}
\usepackage{longtable}
% DOCUMENT LAYOUT
% \usepackage{geometry} 
\usepackage[margin=1.0in]{geometry}
\usepackage{lineno}
\usepackage[usenames,dvipsnames]{xcolor}
\definecolor{myorange}{RGB}{240, 96, 0}
\newcommand{\mt}[1]{{\textcolor{myorange} {({\tiny MT:} #1)}}}
% \geometry{letterpaper, textwidth=5.5in, textheight=8.5in, marginparsep=7pt, marginparwidth=.6in}
% \setlength\parindent{0in}

@@

\newcommand{\sigmapre}{\sigma_\mathrm{pre}}
\newcommand{\sigmapost}{\sigma_\mathrm{post}}
\newcommand{\normalpre}{\mathcal{N}(\mu, \sigmapre)}
\newcommand{\normalpost}{\mathcal{N}(\mu, \sigmapost)}

@@

Our model represnts the generative process of finding different forms of group consensus
in group polarization experiments, and the process of measuring that consensus using
an ordinal measurement scale, e.g.\ a Likert-type scale. This model considers two
relevant forms of consensus: \emph{simple consensus} and \emph{group polarization}.
Either process is \emph{consensus}, if we define consensus as a reduction
in opinion variance over time.  
Group polarization experiments are designed to result group consensus
since they group together like-minded individuals to see if they become more
extreme after discussion~\cite{Turner2018}.

We collected group polarization data from ten empirical studies with sixty constituent
experimental conditions across all studies. We recorded the reported opinion shift and pre- and
post-deliberation mean opinions, and the number of participants
in each condition, as long as this data was available---\citeA{Burnstein1975} 
did not report pre- and post-deliberation
means, only the shift in means. We did not include experimental treatments that did not 
yield a positive detection of group polarization since we are only estimating
the rate of plausible false positive detections of group polarization.

Based on this data

% In two cases, reported opinion and shift values were converted from an ordinal
% scale to a percentage scale, specifically \citeA{Friedkin1999a} and \citeA{Krizan2007}.
% The authors in these papers did report the original ordinal measurement scale,
% so we converted the means back to the orignial measurement scale.
% In a few cases from \citeA{Friedkin1999a} we did not need to apply our 
% model to the reported data because we could determine a reported opinion shift was 
% \emph{prima facie} false since reported opinion shift magnitudes
% were smaller than the standard deviation in shifts (see Table 1, p. 868, \emph{ibid}),
% despite Friedkin's claim that regressions (not clearly defined) on all shifts 
% yielded $p < 0.05$~\cite{Kruschke2018c}.

\subsection{Latent opinions, ordinal measurements, and how simple consensus can look like
group polarization}

Our analysis relies on two important formal concepts: latent opinions and their 
measurement on an ordinal scale. \emph{Latent opinions} are theoretical,
unobservable, psychological, continuous-valued variables that are 
somehow represented and stored in the brain-body, and can be reproduced with
some accuracy when a person is prompted. \emph{Ordinal measurements} of 
of latent opinions are ordinal-valued representations of latent opinions 
in some physical form, such as a mark on a sheet of paper or radio button on a web form, that
indicate the polarity and strength of a person's opinion opinion in terms of
one of a number of opinion bin values.

For example, on a 7-point ordinal scale -3 could indcate 
``strongly disagree'', +3 could be ``strongly agree'', 0 would be psychologically 
neutral, and intermediate values are less extreme opinions of either valence, 
such as was used by \citeA{Moscovici1969}. Or, one might

@@

\subsection{Evaluating whether a group polarization finding is plausibly false}

We used large-$N$ and generative modeling supported by $t$-tests on simulated
data to demonstrate that many published findings are plausibly false.

\subsubsection{Identify null-model parameters that yield plausibly-false positives}

@@

After identifying null-model parameters that yield plausibly-false positives 
in the $N \to \infty$ case, we generated binned samples from normal opinion 
distributions $\normalpre$ and $\normalpost$.

\subsection{Bayesian ordered-probit models reliably identify simple consensus}

@@

\spacing{1.0}
\input{t-test-table.tex}
% \input{t-test-table-custom.tex}
\spacing{1.5}

@@

This problem is made all the worse by the lack of a consistent experimental design for group
polarization experiments. This makes it impossible to systematically account for
additional sources of vairance known to lead to over-estimates
of effect sizes~\cite{Yarkoni2021}, independent of the false positives problem
we identified in the current paper.

% This is due to the rush to explain why 
% group polarization occurs using weak theory~\cite{Cartwright1971,Cartwright1973} and makes
% group polarization meta-analyses (e.g. Brown, 1986; and Isenberg, 1986)\nocite{Brown1986,Isenberg1986}
% unreliable~\cite{Meehl1990,Meehl1997}.
